---
title: "Data Management"
subtitle: "Factors"
output:
  xaringan::moon_reader:
    seal: TRUE
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    css: xaringan-themer.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.retina = 3  ## retina more effective than w/h (according to twitter)
                      # fig.width = 16, 
                      # fig.height = 10
                      )

## I dont know why this is included in the example xaringan slide 
## but is something to do with background images
options(htmltools.dir.version = FALSE)

## install and load necessary packages 
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  tidyverse,  # data management and visualization
  gtsummary,  # summary tables
  kableExtra, # for output tables
  xaringanthemer  # for styling presentation 
)

```


```{r  xaringan-themer, include = FALSE}

## define presentation colours (theme) using {xaringanthemer} package 
## https://pkg.garrickadenbuie.com/xaringanthemer/articles/xaringanthemer.html

## epirhandbook logo colours: 
  ## blue: "#00538c"
  ## green: "#007732"
  ## lighter green: "#48a878"

## see ?style_mono_accent for all the things can customise
style_mono_accent(
  base_color = "#00538c", 
  link_color = "#48a878", 
  ## add logo to the title page (bit bigger)
  title_slide_background_image = "https://github.com/appliedepi/epiRhandbook_eng/raw/master/images/epiRhandbook_HexSticker_500x500.png", 
  title_slide_background_position = "100% 100%", 
  title_slide_background_size = "25%", 
  ## add logo to all following slides
  background_image = "https://github.com/appliedepi/emory_training/raw/master/presentation/images/AppliedEpi_Hex.png", 
  background_size = "10%",
  background_position = "100% 0%"
)

```

```{css, echo=F}
    .remark-slide table{
      border: none
    }

    .remark-slide-table {
      
    }

    tr:first-child {
      border-top: none;
  }
    tr:last-child {
    border-bottom: none;
  }

```
<!-- Slide 1: Introduce the chapter with objectives and a brief outline -->

# Objective
  
1. To describe factors as a class of data, and to demonstrate common use-cases
2. To show how factors are used in R, along with packages that are helpful for working with factors

# Outline
* Introduction to factors in data management
* Converting to factor
* Adding or dropping levels in a factor
* Adjusting factor order

---
<!-- Slide 2: Describe the topic and why it's a needed aspect of data management -->
# [Topic]
#### ["Any over-arching statement on the topic, or plain-language description/definition"]

--

### What is [topic]?
 * [Define the topic in semi-technical language]

--

### How can [topic] facilitate data analysis?
 * [example 1: e.g. "factored data facilitates categorical analysis"]
 * [example 2]

---
<!-- Slide 3: Provide examples of messy data and clean data related to teh topic -->

# [Topic]

<!-- Describe a few examples related to the topic of how data can be messy and of clean data. The idea here is to show how there are many ways data can be messy, but that clean data is more standardized. -->

.pull-left[

<!-- Examples: If using table-based examples, define them here, using simple common data elements like demographics: -->
```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
messy_example <- data.frame(case = c("Case 1", "Case 1", "Case 2", "Case 2", "Case 3", "Case 3"), variable = c("age", "sex", "age", "sex", "age", "sex"), value = c("31", "M", "24", "F", "27", "M"))

clean_example <- messy_example %>%
  pivot_wider(names_from = variable, values_from = value)
```

### Messy data:

<!-- For topics where a table would better describe messy data, replace the bullets with a short table (e.g. 2 columns, 3 or 4 rows) -->
  <!-- * [example 1] -->

  <!-- * [example 2] -->

  <!-- * [example 3] -->

<!-- If using table example, remove the example bullets and print the messy data example with the below: -->

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
     messy_example  %>%
     knitr::kable()
```

]

.pull-right[
### Clean data:
<!-- If using a table for messy data examples, show the same table but cleaned -->

  <!-- * [example 1] -->

<!-- If using table example, remove the example bullets and print the clean data example with the below: -->

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
     clean_example  %>%
     knitr::kable()
```

]

---

<!-- Slide 4: sub-chapter 1 -->

## Sub-chapter heading

[narrative]

### R functions and packages

--

* example text with `functions` and **packages**

```{r, eval=FALSE, echo = TRUE, message=FALSE, warning=FALSE}
#Add r segement to be displayed here
x <- y
```
     
--

* example text with `functions` and **packages**

```{r, eval=FALSE, echo = TRUE, message=FALSE, warning=FALSE}
#Add r segement to be displayed here
y <- z
```
      
---

<!-- Slide 5: sub-chapter 1 -->

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
messy_example <- data.frame(case = c("Case 1", "Case 1", "Case 2", "Case 2"), variable = c("age", "sex", "age", "sex"), value = c("31", "M", "24", "F"))
```
## Sub-chapter heading
#### Messy data input:
```{r, eval=TRUE, echo = FALSE, message=FALSE, warning=FALSE}
     messy_example %>%
          knitr::kable()
```

#### Cleaning code:
```{r, eval=TRUE, echo = TRUE, message=FALSE, warning=FALSE}
     clean_example <- messy_example %>%
          pivot_wider(names_from = variable, values_from = value)
```

#### Clean data output:
```{r, eval=TRUE, echo = FALSE, message=FALSE, warning=FALSE}
     clean_example %>%
          knitr::kable()
```
---

# In-depth example (if needed)

.pull-left[

```{r, eval=F, echo=T}
clean_example <- messy_example %>%
     # step1
```


Description of the above code with `functions` and **packages**

]

.pull-right[

```{r, eval=T, echo=F}
  messy_example %>%
     # step1
     knitr::kable()
```

]



---

# In-depth example (if needed)

.pull-left[

```{r, eval=F, echo=T}
clean_example <- messy_example %>%
     # step1 
*    # step2
```


Description of the above code with `functions` and **packages**

]

.pull-right[

```{r, eval=T, echo=F}
  messy_example %>%
     # step1 %>%
     # step2 %>%     
     knitr::kable()
```

]



---

# In-depth example (if needed)

.pull-left[

```{r, eval=F, echo=T}
clean_example <- messy_example %>%
     # step1 
     # step2
*    # step3
```


Description of the above code with `functions` and **packages**

]

.pull-right[

```{r, eval=T, echo=F}
  messy_example %>%
     # step1 %>%
     # step2 %>%
     # step3 %>%     
     knitr::kable()
```

]

---

