---
title: "Situation Report"
subtitle: "Ebola outbreak in Sierra Leone"
author: "(Your name or agency here)"
output:
  word_document: default
date: "`r Sys.Date()`"
params:
  district: "Central Hospital"
  publish_date: "2014-12-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r}
# Install packages
pacman::p_load(
     rio,
     here,
     janitor,
     epikit,
     apyramid,
     lubridate,
     gtsummary,
     scales,
     gghighlight,
     ggExtra,
     flextable,
     tidyverse
)
```


```{r}
# Import each dataset

# surveillance
surv_raw <- import(here("data", "raw", "surveillance_linelist_12012014.csv"))

# hospital datasets
hosp_central  <- import(here("intro_course", "data", "raw", "hospitals", "20141201_hosp_central.csv"))
hosp_military <- import(here("intro_course", "data", "raw", "hospitals", "20141201_hosp_military.csv"))
hosp_other    <- import(here("intro_course", "data", "raw", "hospitals", "20141201_hosp_other.csv"))
hosp_port     <- import(here("intro_course", "data", "raw", "hospitals", "20141201_hosp_port.csv"))
hosp_smmh     <- import(here("intro_course", "data", "raw", "hospitals", "20141201_hosp_smmh.csv"))
hosp_missing  <- import(here("intro_course", "data", "raw", "hospitals", "20141201_hosp_missing.csv"))

# laboratory dataset
lab <- import(here("intro_course", "data", "raw", "lab_results_20141201.xlsx"))

# case investigation dataset
investigations <- import(here("intro_course", "data", "raw", "case_investigations_20141201.xlsx"))
```



```{r}
# Clean the surveillance linelist
surv <- surv_raw %>% 
  
  # automatically clean column names
  clean_names() %>% 
  
  # manually clean column names   
  rename(
    date_onset = onset_date,
    date_report = date_of_report,
    district_res = adm3_name_res,
    district_det = adm3_name_det) %>%
     
  # remove unnecessary column
  select(-row_num) %>% 

  # de-duplicate rows  
  distinct() %>% 
     
  # convert date_onset to date class
  mutate(date_onset = mdy(date_onset)) %>% 
  mutate(date_report = mdy(date_report)) %>% 
     
  # create epiweek columns  
  mutate(week_onset = floor_date(date_onset, unit = "week")) %>% 
  mutate(week_report = floor_date(date_report, unit = "week")) %>% 
     
  # convert age to numeric class
  mutate(age = as.numeric(age)) %>% 
     
  # properly record missing values
  mutate(across(.cols = where(is.character), .fns = na_if, "")) %>% 
     
  # Make date-difference column  
  mutate(diff = date_report - date_onset) %>% 
     
  # convert negative values to NA
  mutate(wt_kg = ifelse(wt_kg < 0, NA, wt_kg),
         bmi   = ifelse(bmi < 0,   NA, bmi)) %>% 
     
  # convert gender values to full words
  mutate(gender = case_when(               # re-define gender as: 
    gender == 'm' ~ 'male',                # when "m", change to "male"   
    gender == 'f' ~ 'female',              # when "f", change to "female" 
    TRUE          ~ gender)) %>%           # any other value, remain as before
     
  # create age-in-years
  mutate(age_years = case_when(
    age_unit == "years"  ~ age,            # if age is given in years
    age_unit == "months" ~ age/12,         # if age is given in months
    is.na(age_unit)      ~ age,            # if age unit is missing, assume years
    TRUE                 ~ NA_real_)) %>%  # any other circumstance, assign missing
       
  # create age category column
  mutate(age_cat = age_categories(         # create new column
    age_years,                             # numeric column to make groups from
    lower = 0,
    upper = 70,
    by = 10)) %>% 
     
  # create column marking TRUE if district of residence and detection differ
  mutate(moved = district_res != district_det) %>% 
  
  # create new column that prioritizes district of detection
  mutate(district = coalesce(district_det, district_res)) %>% 
  
  # re-code hospital column
  mutate(hospital = recode(hospital,
    # for reference: OLD = NEW
    "Mitilary Hospital"  = "Military Hospital",
    "Port"               = "Port Hospital",
    "Port Hopital"       = "Port Hospital",
    "St. Mark's Maternity Hospital (SMMH)" = "SMMH")) %>% 
     
  # remove suspect cases
  filter(case_def == "Confirmed")
```


```{r}
# bind the rows of the hospital linelists
hosp <- bind_rows(hosp_central, hosp_port, hosp_military, hosp_smmh, hosp_other, hosp_missing)
```

```{r}
# Modify the hosp dataset
hosp <- hosp %>% 
  # select specific columns from hosp, and re-name ID as case_ID
  select(
    case_id = ID,               # select and rename
    `hospitalisation date`,     # select
    `admission time`,           # select
    `outcome date`,             # select
    outcome)                    # select

# Join the two data frames with a full-join
combined <- full_join(surv, hosp, by = "case_id")
```

```{r}
# Join the surveillance and hospital data frames with a full-join
# (place this in the Joining data section of your script)
combined <- full_join(combined, lab, by = "case_id")

```

```{r}
# Case investigations dataset
  # keep only certain columns  
  # (add to the Joining data section of your R script)
investigations <- investigations %>% 
  select(-c(age, `age unit`, gender))

# Join the two data frames with a full-join
combined <- full_join(combined, investigations, by = "case_id")
```



```{r}
# Clean the new columns that have been joined to 'combined'
combined <- combined %>% 
  
  # convert all column names to lower case and remove spaces
  clean_names() %>% 
  
  # edit names of new date columns
  rename(date_hospitalisation = hospitalisation_date,
         date_outcome         = outcome_date,
         date_infection       = infection_date) %>% 
  
  # covert new columns to class date
  mutate(date_hospitalisation = mdy(date_hospitalisation),
         date_outcome         = mdy(date_outcome),
         date_infection       = ymd(date_infection))
```


```{r}
# Define data date as Sunday of prior week
data_date <- floor_date(ymd(params$publish_date)-7, unit = "week")
```

```{r}
# Filter combined dataset to data_date
combined <- combined %>% 
  filter(date_report <= data_date)
```



```{r}
export(combined, here("data", "linelist_combined_20141201.rds"))
```


# Executive summary

The outbreak of Ebola continues, although reported cases are decreasing in all districts following the surge in contract-tracing support provided by international partners.

"As of `r format(data_date, "%d %B, %Y")` there had been `r nrow(combined)` confirmed cases reported."  


| Partner                                 | Arrived  | Activities                   |
|-----------------------------------------|----------|------------------------------|
| Doctors without Borders (MSF) - Belgium | 20 July  | Treatment centers            |
| World Health Organization (WHO)         | 22 July  | Surveillance and analytics   |
| Red Cross (ICRC)                        | 1 August | Community-based surveillance |

: Table of partners involved in the response



## Person

```{r}
# Age pyramid with ggplot modifications
age_pyramid(
     data = combined,
     age_group = "age_cat",
     split_by = "gender",
     proportional = TRUE,
     show_midpoint = FALSE)+
     
  theme_minimal()+
     
  scale_fill_brewer(type = "qual", palette = 2)+
  
  labs(title = str_glue("Age and gender of confirmed cases"),
       x = "Proportion of all cases",
       y = "Age group",
       caption = str_glue("Ebola outbreak"),
       fill = "Gender")
```

## Time

```{r}

# define weekly breaks from Monday before first case

# Define and save the vector
ebola_weeks <- seq.Date(
  from = floor_date(min(combined$date_onset, na.rm=T), unit = "week", week_start = 1),
  to =   ceiling_date(max(combined$date_onset, na.rm=T), unit = "week", week_start = 1),
  by =   "week")


# Run the plot, using the vector
ggplot(data = combined, aes(x = date_onset)) +
  geom_histogram(breaks = ebola_weeks, closed = "left") +
  labs(x = "Date of onset", y = "Incidence") +
  theme_bw()


# Make epidemic curve
ggplot(data = combined)+
       geom_histogram(
          mapping = aes(x = date_onset),
          breaks = ebola_weeks,
          closed = "left")+
     
       # x-axis labels
       scale_x_date(
         expand            = c(0,0),           # remove excess x-axis space before and after case bars
         date_breaks       = "4 weeks",        # date labels and major vertical gridlines appear every 3 Monday weeks
         date_minor_breaks = "week",           # minor vertical lines appear every Monday week
         labels = scales::label_date_short())+ # automatically efficient date labels
       
       # y-axis
       scale_y_continuous(
         expand = c(0,0))+             # remove excess y-axis space below 0 (align histogram flush with x-axis)
       
       # aesthetic themes
       theme_minimal()+                # simplify plot background
       
       theme(
         plot.caption = element_text(hjust = 0,        # caption on left side
                                     face = "italic"), # caption in italics
         axis.title = element_text(face = "bold"))+    # axis titles in bold
       
       # labels including dynamic caption
       labs(
         title    = "Weekly incidence of cases (Monday weeks)",
         subtitle = "",
         x        = "Week of symptom onset",
         y        = "Weekly incident cases reported",
         caption  = stringr::str_glue("n = {nrow(combined)} confirmed cases; Onsets range from {format(min(combined$date_onset, na.rm=T), format = '%a %d %b %Y')} to {format(max(combined$date_onset, na.rm=T), format = '%a %d %b %Y')}\n{nrow(combined %>% filter(is.na(date_onset)))} missing date of onset and not shown"))
```

## Place

```{r}
combined %>% 
     mutate(district = fct_explicit_na(district, na_level = "Missing")) %>% 
     tabyl(district) %>% 
     arrange(desc(n)) %>% 
     mutate(percent = scales::percent(percent, 1)) %>% 
     qflextable()
```
