---
title: "Installing R...TODO "
subtitle: "Subtitle here"
output: 
  html_document:
    toc: true
    toc_depth: 2
theme: sandstone
date: 'Date revised: 2022-03-02'
---


```{r, echo=FALSE, include = F}
# Put logo in upper-right of document
htmltools::img(src = knitr::image_uri(here::here("images", "logo.png")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;',
               width = "200px",
               heigth = "200px")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview


# Preparation  

Contact IT

# Permissions and Network drives

# R

# RStudio

# R Packages

Instructions here on how to run this script. Note that the students may have no idea what an R package is. If they encounter any errors running the script or don't know what to do on their own, don't worry, just email the course instructor and we'll sort it out in a one-on-one troubleshooting session. 


```{r, message=FALSE, warning=FALSE}
#############################################################
# Essential R packages: Intro to R for Applied Epidemiology #
#############################################################

# This script uses the p_load() function from pacman R package, 
# which installs if package is absent, and loads for use if already installed


# Ensures the package "pacman" is installed
if (!require("pacman")) install.packages("pacman")


# Packages available from CRAN
##############################
pacman::p_load(
     
     # learning R
     ############
     learnr,   # interactive tutorials in RStudio Tutorial pane

     # project and file management
     #############################
     here,     # file paths relative to R project root folder
     rio,      # import/export of many types of data
     openxlsx, # import/export of multi-sheet Excel workbooks 
     
     # package install and management
     ################################
     pacman,   # package install/load
     remotes,  # install from github
     
     # General data management
     #########################
     tidyverse,    # includes many packages for tidy data wrangling and presentation
          #dplyr,      # data management
          #tidyr,      # data management
          #ggplot2,    # data visualization
          #stringr,    # work with strings and characters
          #forcats,    # work with factors 
          #lubridate,  # work with dates
          #purrr       # iteration and working with lists
     epitrix,      # Useful epi functions

     
     # statistics  
     ############
     janitor,      # tables and data cleaning
     gtsummary,    # making descriptive and statistical tables
     rstatix,      # quickly run statistical tests and summaries

     # plots - general
     #################
     #ggplot2,         # included in tidyverse
     cowplot,          # combining plots  
     RColorBrewer,     # color scales
     incidence2,   # Make epicurves and handle incidence data
     i2extras,     # Extra functions for the incidence2 package
     gghighlight,      # highlight a subset
     ggrepel,          # smart labels

     # routine reports
     #################
     rmarkdown,        # produce PDFs, Word Documents, Powerpoints, and HTML files
     reportfactory,    # auto-organization of R Markdown outputs
     officer,          # powerpoints
     
     # dashboards
     ############
     flexdashboard,    # convert an R Markdown script into a dashboard
  
     # tables for presentation
     #########################
     knitr,            # R Markdown report generation and html tables
     flextable        # HTML tables
     #DT,              # HTML tables (alternative)
     #gt,              # HTML tables (alternative)
     #huxtable,        # HTML tables (alternative) 

)

```

