---
title: "Data Management"
subtitle: "Cleaning data and core functions"
output:
  xaringan::moon_reader:
    seal: TRUE
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    css: xaringan-themer.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.retina = 3  ## retina more effective than w/h (according to twitter)
                      # fig.width = 16, 
                      # fig.height = 10
                      )

## I dont know why this is included in the example xaringan slide 
## but is something to do with background images
options(htmltools.dir.version = FALSE)

## install and load necessary packages 
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  tidyverse,  # data management and visualization
  gtsummary,  # summary tables
  kableExtra, # for output tables
  xaringanthemer  # for styling presentation 
)

```


```{r  xaringan-themer, include = FALSE}

## define presentation colours (theme) using {xaringanthemer} package 
## https://pkg.garrickadenbuie.com/xaringanthemer/articles/xaringanthemer.html

## epirhandbook logo colours: 
  ## blue: "#00538c"
  ## green: "#007732"
  ## lighter green: "#48a878"

## see ?style_mono_accent for all the things can customise
style_mono_accent(
  base_color = "#00538c", 
  link_color = "#48a878", 
  ## add logo to the title page (bit bigger)
  title_slide_background_image = "https://github.com/appliedepi/epiRhandbook_eng/raw/master/images/epiRhandbook_HexSticker_500x500.png", 
  title_slide_background_position = "100% 100%", 
  title_slide_background_size = "25%", 
  ## add logo to all following slides
  background_image = "https://github.com/appliedepi/emory_training/raw/master/presentation/images/AppliedEpi_Hex.png", 
  background_size = "10%",
  background_position = "100% 0%"
)

```

```{css, echo=F}
    .remark-slide table{
      border: none
    }

    .remark-slide-table {
      
    }

    tr:first-child {
      border-top: none;
  }
    tr:last-child {
    border-bottom: none;
  }

```

# Objective
  
1. To demonstrate common steps and best practices used in the process of cleaning a dataset
2. To introduce essential R packages and functions used in data cleaning

# Outline
1. Characteristics of clean data
2. Steps in data cleaning  
     * Importing data
     * Structuring a dataset
     * Formatting data
     * Trimming a dataset
     * Coding values
     * Arranging and sorting a dataset     

---
# Characteristics of clean data

.pull-left[

- Goal of data cleaning:
  - Structure and format data in a way that facilitates data analysis
- Characteristics of clean data structure:  
  - Each variable must have its own column
  - Each observation must have its own row
  - Each value must have its own cell

]

.pull-right[

Illustrations of each characteristic of clean data to be added here

]

---
class: center, middle

.footnote[
Slides created via the [{xaringan}](https://github.com/yihui/xaringan) R package
]