---
title: "Data Management"
subtitle: "Cleaning data and core functions"
output:
  xaringan::moon_reader:
    seal: TRUE
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    css: xaringan-themer.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.retina = 3  ## retina more effective than w/h (according to twitter)
                      # fig.width = 16, 
                      # fig.height = 10
                      )

## I dont know why this is included in the example xaringan slide 
## but is something to do with background images
options(htmltools.dir.version = FALSE)

## install and load necessary packages 
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways  
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  tidyverse,  # data management and visualization
  gtsummary,  # summary tables
  kableExtra, # for output tables
  xaringanthemer  # for styling presentation 
)

```


```{r  xaringan-themer, include = FALSE}

## define presentation colours (theme) using {xaringanthemer} package 
## https://pkg.garrickadenbuie.com/xaringanthemer/articles/xaringanthemer.html

## epirhandbook logo colours: 
  ## blue: "#00538c"
  ## green: "#007732"
  ## lighter green: "#48a878"

## see ?style_mono_accent for all the things can customise
style_mono_accent(
  base_color = "#00538c", 
  link_color = "#48a878", 
  ## add logo to the title page (bit bigger)
  title_slide_background_image = "https://github.com/appliedepi/epiRhandbook_eng/raw/master/images/epiRhandbook_HexSticker_500x500.png", 
  title_slide_background_position = "100% 100%", 
  title_slide_background_size = "25%", 
  ## add logo to all following slides
  background_image = "https://github.com/appliedepi/emory_training/raw/master/presentation/images/AppliedEpi_Hex.png", 
  background_size = "10%",
  background_position = "100% 0%"
)

```

```{css, echo=F}
    .remark-slide table{
      border: none
    }

    .remark-slide-table {
      
    }

    tr:first-child {
      border-top: none;
  }
    tr:last-child {
    border-bottom: none;
  }

```

# Objective
  
1. To demonstrate common steps and best practices used in the process of cleaning a dataset
2. To introduce essential R packages and functions used in data cleaning

# Outline
* Characteristics of clean data
* Steps in data cleaning  
     * Importing data
     * Naming and selecting columns
     * Deduplication
     * Column creation, transformation, and re-coding
     * Adding and filtering rows
     * Arranging and sorting a dataset     
     
---
# Characteristics of clean data

#### Clean data is data that has been structured and formated in a way that facilitates data analysis

### Characteristics of clean data structure:  
[Note: This and the following three slides could be moved to the pivoting data slide module]
--

1. Each variable must have its own column

--

2. Each observation must have its own row

--

3. Each value must have its own cell

---
```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

clean_data_example_1 <- data.frame(case = c("Case 1", "Case 1", "Case 2", "Case 2", "Case 3", "Case 3"), variable = c("age", "sex", "age", "sex", "age", "sex"), value = c("31", "M", "24", "F", "27", "M")) 

clean_data_example_2 <- clean_data_example_1 %>%
  pivot_wider(names_from = variable, values_from = value) 
```

### 1.  Each variable must have its own column

.pull-left[
#### Messy:
```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

clean_data_example_1  %>%
  knitr::kable()
  
```
]

.pull-right[
#### Clean:
```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

clean_data_example_2  %>%
  knitr::kable()

```
]

---
```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

clean_data_example_3 <- data.frame(clinic = c("Clinic1", "Clinic2", "Clinic3"), cases_2017 = c(19, 65, 72), cases_2018 = c(8, 85, 32), cases_2019 = c(12, 45, 88)) 

clean_data_example_4 <- clean_data_example_3 %>%
  pivot_longer(cols=2:4, names_to="year", values_to = "cases") %>%
  mutate(year = as.numeric(substr(year, 7,10))) 
```

### 2.  Each observation must have its own row

.pull-left[
#### Messy:
```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

clean_data_example_3  %>%
  knitr::kable()
  
```
]

.pull-right[
#### Clean:
```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

clean_data_example_4  %>%
  knitr::kable()

```
]

---
```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

clean_data_example_5 <- data.frame(case = c("Case 1", "Case 2", "Case 3", "Case 4"), age_sex = c("31, M", "24, F", "24, M", "33, F")) 

clean_data_example_6 <- clean_data_example_5 %>%
    separate(age_sex, 
           into = c("age", "sex"), 
           sep = ", "
           ) %>%
  mutate(age = as.numeric(age))
```

### 3. Each value must have its own cell

.pull-left[
#### Messy:
```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

clean_data_example_5  %>%
  knitr::kable()
  
```
]

.pull-right[
#### Clean:
```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

clean_data_example_6  %>%
  knitr::kable()

```
]

.pull-left[

]
---
# Steps in Cleaning Data

### Importing data

[This slide will cover "8.3 Import data"]

---
# Steps in Cleaning Data

### Name and Select Columns

[This slide will cover "8.4 Column names" and 8.5 Select or re-order columns"]

---
# Steps in Cleaning Data

### Deduplication

[This slide will cover "8.6 Deduplication"]

---
# Steps in Cleaning Data

### Column creation, transformation, and re-coding

[This slide will cover "8.7 Column creation and transformation", "8.8 Re-code values", and "8.9 Numeric categories"]

---
# Steps in Cleaning Data

### Adding and filtering rows

[This slide will cover "8.10 Add rows" and "8.11 filter rows", and "8.12 row-wise calculations"]

---
# Steps in Cleaning Data

### Arranging and sorting a dataset

[This slide will cover "8.12 Arrange and sort"]
