---
title: "COVID-19 case study tip sheet"
subtitle: ""
output: html_document
theme: sandstone
date: 'Date revised: 2022-06-09'
params:
  data_date: "2021-06-21"
---


```{r, echo=FALSE}
# Put logo in upper-right of document
htmltools::img(src = knitr::image_uri(here::here("images", "logo.png")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;',
               width = "200px",
               heigth = "200px")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=F}
pacman::p_load(rio, here, epikit, lubridate, tidyverse)
combined <- import(here("intro_course", "data", "clean", "backup", "linelist_combined_20141201.rds"))
```

In this sheet, you will find some hints and tips for re-creating the COVID-19 situation report.  

## YAML

Begin with the standard YAML. Recognize that the report is as of 21 June 2021, so you need to specify that date as a parameter called "data_date" (or another name) that you can use to filter the data throughout the report.  

The easiest method is to simply add a `params` section to your YAML and add this date as a character value, as below. YAML is very specific about spaces, colons, and indentations.  

`````
---
params:
  datadate: "21 June 2021"
---
`````

But this method is not the most easy for colleagues to edit. You can make the date selection by point-and-click interface using the below setup.  
Now, if you click to "Knit with parameters", a pop-up window will ask you to select the `data_date`.

`````
---
params:
  data_date:
    label: "Date of data:"
    value: "2021-06-21"
    input: date
---
`````

**Either way** - you must reference this parameter in your R code as `params$data_date`, and it will be class *character*. So typically you will reference it as `ymd(params$data_date)`, if for example you are using it to filter cases.  

```{r, echo=TRUE}
ymd(params$data_date)
class(ymd(params$data_date))
```



## Inline text summary at top of report  

- Think about order of your code. Don't forget that in order to print these text statement with summary statistics inline, those calculations and data cleaning need to happen *above* this text in the Rmd (but not emit any visible output).  

- To display dates in a specific format, you can wrap the date in `format()` and supply the `strptime` syntax that you prefer, for example:

```{r, echo=T, eval=T}
format(ymd(params$data_date), format = "%d %B, %Y")
```

- To display counts and (%) quickly, you can use the `fmt_count()` function from {epikit}. This accepts the name of a data frame, and then a logical statement used to filter rows.  

```{r, echo=T, eval=T}
fmt_count(combined, outcome == "Death")
```


## Page breaks  

To get page breaks when you want them, insert **\\pagebreak** into the Rmd  


## Data cleaning

- You can begin by using the `clean_names()` function from {janitor} to convert all column names to lowercase with no spaces or special characters  

- Then you can use `rename()` - REMEMBER the expected syntax is **NEW = OLD**  

- Convert the important date columns to class "date" using one of the {lubridate} functions, `dmy()`, `mdy()`, or `ymd()`. **Use the function that corresponds to the how the dates are formatted  prior to applying the function.**  

- The report is for *confirmed cases* before 21 June 2021, so do not forget to apply those criteria using `filter()`.  




## Age pyramid  

- Use the {apyramid} package and its `age_pyramid()` function (don't forget to add {apyramid} to your packages!). This creates a ggplot object, so you can modify it by adding a `+` afterward and then use `labs()` to add labels like title and caption. Review the epidemiological plots exercise if you do not remember this function, or search for the package documentation.  

## Weekly epidemic curve, by city  

- To break the histogram into weeks, you can either set `binwidth = 7`, or to be more precise you can define a series of weekly break values that start the Monday prior to the first case, and end the Monday after the last case.  

```{r, echo=T, eval=F}
weekly_breaks <- seq.Date(
      from = floor_date(min(linelist$date_report, na.rm=T),   "week", week_start = 1), # Monday before first case
      to   = ceiling_date(max(linelist$date_report, na.rm=T), "week", week_start = 1), # Monday after last case
      by   = "week")                                                                   # bins are 7-days 

```

- The you need to use these breaks within `geom_histogram()`.  

- To make the dates appear nicely efficient (no duplicate years), recall how to assign `label_date_short()` to the `labels = ` argument of `geom_histogram()`.  

- Observe how the list of cities (displayed, and in the legend) has been truncated to 5 and "Other"... this was done using `fct_lump_n()` within the ggplot command, for example:  


```{r, echo=T, eval=F}
ggplot(data = linelist, mapping = aes(x = report_dt, fill = fct_lump_n(city, 5)))+
     geom....
```


## Race and outcome  

- This table is produced with {gtsummary} and it's function `tbl_summary()`. Add p-values by piping into `add_p()`. Don't forget to only select the columns you need before sending the data into `tbl_summary()`.  

## Cumulative case incidence table  

- You will need to first use `group_by()` by city and then `summarise()` the number of cases per city.  
- You add other columns using `sum()` and logical statements within (don't forget `na.rm = T`).  
  - For example, to calculate cases reported in the 14 days prior to a `data_date` parameter, you could include this within the `summarise()`:  
  - `recent_14d = sum(date_report >= ymd(params$data_date) - 14)`  
- The final column for cumulative incidence is created by joining the summarized table to the data frame `city_pops`. You want to keep all the rows in the summarized table, but only the rows in `city_pops` that match.  
- You then have the population data to calculate the number of cumulative cases per population. Use `(total_cases / population) * 10000` to calculate the incidence rate per 10000 population.  
- You can use `round()` with `digits = 1` to get the rounding on the cumulative incidence.  


## General

Review your `setup` chunk near the top and set `echo = FALSE` to not show the code in the output.  



## Good luck!  

Just do the best you can. This is not a test. It is a challenging exercise, but part of the learning process. 
     