---
title: "Covid-19 outbreak situation report"
author: "Case study exercise for Applied Epi R course"
date: "30 June 2021"
output:
  word_document: default
  html_document: default
params:
  surveillance_date: '2021-06-30'
---


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
This comment will not show up when you knit the document.
A comment with a title with slashes indicates a name of a code chunk.
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// setup \\\
--------------------------------------------------------------------------------

Several packages are required for different aspects of analysis with *R*. 
You will need to install these before starting. 

We install and load packages using the {pacman} package. Its p_load() command
will install packages if necessary and load them for use in the current session.

This might prove difficult if you have limited administrative rights for your 
computer. Making sure your IT-department gives you the correct access can save a 
lot of headache. 

See this handbook pages on the basics of installing packages and running R from 
network drives (company computers) for more detail. 

https://epirhandbook.com/r-basics.html#installation
https://epirhandbook.com/r-on-network-drives.html#r-on-network-drives
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->


```{r setup, include=FALSE}

# hide all code chunks in the output, but show errors 
     # nb. warnings and messages still shown
knitr::opts_chunk$set(echo = FALSE,  # hide all code chunks in output
                      error = TRUE,  # show errors if they appear, but don't stop (produce the word doc)
                      warning = FALSE, # do not show warnings in the output word doc 
                      message = FALSE, # do not show  messages in the output word doc
                      fig.width = 7,         # Figure width
                      fig.height = 7,        # Figure height
                      fig.topcaption = TRUE  # show figure titles on top of plot
                     )

# Ensures the package "pacman" is installed
if (!require("pacman")) {
     install.packages("pacman") }

# install (if necessary) from CRAN and load packages to be used
pacman::p_load(
  rio,        # importing data  
  here,       # relative file pathways 
  skimr,      # review data
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  epikit,     # age_categories() function
  tidyverse,  # data management and visualization
  flextable,  # converting tables to pretty images
  scales,     # define colour schemes for flextables 
  gtsummary,  # summary statistics, tests and regressions 
  apyramid    # age pyramids 
)


```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// define_reporting_periods \\\
--------------------------------------------------------------------------------
Here we use the date you want to report on, defined in parameters (params) at
top of script to create a date object for filtering datasets, plots/tables, etc. 

We also create a week object to make grouping data easier. In this scenario we 
have defined the week to start on Mondays (day 1 of the week) 

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r define_reporting_periods}

# create a date object from the parameter
# Minus 7 days surveillance date to account for lag in reporting lab results
surveillance_date <- ymd(params$surveillance_date) - 7

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// import \\\
--------------------------------------------------------------------------------

We use the {rio} package for importing our example data - it is very versatile 
and can read in most file types. 

We use the {here} package for defining the path to our file. This is important 
for sharing your script with others (by email or on Sharepoint) - if you used an
"absolute" path, they would need to update the script to match their computer. 

This way your whole R-project folder can be zipped up and moved somewhere else. 

For more details see: 
https://epirhandbook.com/import-and-export.html
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r import}
  
# import the raw case data set
  # define the path using {here} then pass that to the {rio} import function
  # specify the sheet to read using which (default is to read first sheet)
linelist_raw <- rio::import(
  file = here::here("covid_case_study", "covid_example_data.xlsx"))

city_pops <- rio::import(
  file = here::here("covid_case_study", "city_pops.xlsx"))

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// browse_data \\\
--------------------------------------------------------------------------------

Note that this chunk is set to 'eval = FALSE' so the results do not show.  

Here we take a look at the raw data to get a feel for what needs cleaning. 

We first use the in-built browser with the {base} function View(). 

Then we can use the {base} function summary(), but probably the most comprehensive 
overview is with the {skimr} function. 
You can also view distinct values for variables using the {base} unique() function. 

For more details see: 
https://epirhandbook.com/cleaning-data-and-core-functions.html#review
https://epirhandbook.com/descriptive-tables.html#browse-data
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r browse_data, eval = FALSE}

# view your whole dataset interactively (in an excel style format)
View(linelist_raw)

# get summary: 
# mean, median and max values of numeric variables
# counts for categorical variables
# also gives number of NAs
summary(linelist_raw)

# get information about each variable in a dataset 
# nb. “POSIXct” is a type of raw date class 
skim(linelist_raw)

# view unique values contained in variables - useful for categorical variables
# you can run this for any column -- just replace the column name
unique(linelist_raw$case_gender) 


# we can also loop through all categorical variables to look at all possibilities
# see later in the script for introductions to using iteration
purrr::map(
  # all column names (inputs)
  .x = linelist_raw %>% 
    select(where(is_character)) %>% 
    names(), 
  # applies select() then unique() to each column
  .f = ~select(linelist_raw, .x) %>% 
    unique()
  )
```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// clean_colnames \\\
--------------------------------------------------------------------------------

Here we are going to clean the column names of our data set - and store as a new
dataset called "linelist". 

It is possible to use the {janitor} package for automated cleaning of variable 
names - but as there are only a few variables that we want to rename, 
here we will demonstrate using {dplyr} select() function for manually renaming.

Select() can be used either to retain specific columns or to rename them by using
the syntax New name = Old name. 

For more details see: 
https://epirhandbook.com/cleaning-data-and-core-functions.html#column-names
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r clean_colnames}

# create a new object called linelist and assign linelist_raw with renamed columns
linelist <- linelist_raw %>% 
  # use select() to retain columns and rename them 
     # NEW name = OLD name
     # aligned for readability
  select( 
    pid                 = PID,
    date_report         = reprt_creationdt_FALSE,      
    age                 = case_age,                    
    gender              = case_gender,
    race                = case_race,
    eth                 = case_eth,
    city                = City,
    zip                 = case_zip,
    # county              = case_county,
    # district            = case_district,
    # state               = case_state,
    contact_id          = Contact_id, 
    date_onset          = sym_startdt_FALSE,
    sym_fever,
    sym_subjfever,
    sym_myalgia,
    sym_losstastesmell,
    sym_sorethroat,
    sym_cough,
    sym_headache,
    sym_resolved,
    date_recovery       = sym_resolveddt_FALSE, 
    contact_hh          = contact_household,
    hospitalized,  
    date_hospitalized   = hosp_admidt_FALSE,
    date_discharge      = hosp_dischdt_FALSE,
    died,  
    died_covid,  
    date_died           = died_dt_FALSE,
    confirmed_case, 
    covid_dx, 
    date_positive       = pos_sampledt_FALSE,
    lat                 = latitude_JITT,
    lon                 = longitude_JITT
    )

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// clean_dates \\\
--------------------------------------------------------------------------------

Here we are going to clean the date variables 

In the class we have taught using the {lubridate} functions ymd(), mdy(), and dmy(), which also should correspond to the format in the raw data.

The {base} way to convert columns to Date class is with as.Date(). If the
current format is YYYY-MM-DD or YYYY/MM/DD then no other arguments are needed. 
If the format is different, specify it to the argument format= (for more detail see: 
https://epirhandbook.com/working-with-dates.html#convert-to-date )


~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r clean_dates}

linelist <- linelist %>% 
  # convert all date columns to date type 
  # note that dates must be formatted correctly - in some cases date parsers from lubridate can be used here
  mutate(
    date_report =       ymd(date_report),
    date_onset =        ymd(date_onset),
    date_hospitalized = ymd(date_hospitalized),
  ) %>% 
  
     
  # remove onset, hospitalisation, and death dates prior to 2020
  # mutate(
  #   date_onset        = replace(date_onset, date_onset < ymd("2020-01-01"), NA),
  #   date_hospitalized = replace(date_hospitalized, date_hospitalized < ymd("2020-01-01"), NA)
  #   ) %>%
       
     
     
  # remove cases reported after the surveillance_date (for this report)
  filter(date_report <= surveillance_date)  
  

  # create an "epiweek" column from the report date 
  # floor_date() rounds down to the closest week
  # mutate(epiweek = floor_date(date_report,
  #                         # round by weeks
  #                         unit = "week", 
  #                         # define week to start on Sunday
  #                         week_start = 7))
  #

```


<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// clean_categories \\\
--------------------------------------------------------------------------------

create age groups based on the age variable using 
the age_categories() function from {epikit}. It is also possible to create 
age groups using the {dplyr} case_when() function - but is more involved. 

For more details see: 
https://epirhandbook.com/cleaning-data-and-core-functions.html#column-creation-and-transformation
https://epirhandbook.com/cleaning-data-and-core-functions.html#re-code-values
https://epirhandbook.com/characters-and-strings.html?q=regex#regular-expressions-regex
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r clean_categories}

linelist <- linelist %>% 
     
     # create age group variable
     mutate(
       age_group = age_categories(age,
        # define break points
        lower = 0,
        upper = 70,
        by    = 10)) %>% 

     # convert "Unknown" to NA in the column gender
     mutate(gender = na_if(gender, "Unknown")) %>% 

     # convert the race values to title case 
     mutate(race = str_to_title(race))
```



# Summary 

- As of `r format(surveillance_date, format = "%B %d %Y")`, Fulton County has 
recorded `r nrow(linelist)` confirmed cases of COVID-19.  

- As of `r format(surveillance_date, format = "%B %d %Y")`, Fulton County has 
recorded `r fmt_count(linelist, died == "Yes")` deaths.  

- Among all confirmed cases of COVID-19 in Fulton County, `r fmt_count(linelist, hospitalized == "Yes")`
required hospitalization. 


\pagebreak 

# Demographics

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// age_pyramid \\\
--------------------------------------------------------------------------------

Here we demonstrate how to plot age pyramids using the {apyramid} package and edit
it with {ggplot2]. 

For more details see: 
https://epirhandbook.com/demographic-pyramids-and-likert-scales.html
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r age_pyramid}
# plot age pyramid 
linelist %>% 
  
 age_pyramid(
  age_group = "age_group",
  split_by = "gender",
  
  # Show as percentages of total cases
  proportional = TRUE,
  
  # remove guide line for mid-point
  show_midpoint = FALSE) +
  
  # set theme to basic 
  theme_minimal() +
  
  # add labels 
  labs(
    title = "Age and gender of COVID-19 cases in Fulton County",
    subtitle = ,
    caption = "Fake dataset for course exercise",
    x = "Age group",
    y = "Percent of total",
    fill = "Gender")

```



\pagebreak 

# Weekly epidemic curve by city  

```{r epicurve}
# specify weekly breaks on Mondays
weekly_breaks <- seq.Date(
      from = floor_date(min(linelist$date_report, na.rm=T),   "week", week_start = 1), # Monday before first case
      to   = ceiling_date(max(linelist$date_report, na.rm=T), "week", week_start = 1), # Monday after last case
      by   = "week")    # bins are 7-days 


# open a plot with the linelist data
ggplot(linelist, mapping = aes(x = date_report, fill = fct_lump_n(city, 5))) +
  
  geom_histogram(
       breaks = weekly_breaks,  # use breaks defined above
       closed = "left"          # count cases from the start of a breakpoint
  ) +
  
  scale_y_continuous(
    # make the y axis start at the origin
    expand = c(0, 0))+
  
  scale_x_date(
    expand            = c(0,0),            # remove excess x-axis space before and after case bars
    date_breaks       = "2 months",        # date labels and major vertical gridlines appear every 4 Monday weeks
    #date_minor_breaks = "week",           # minor vertical lines appear every Monday week
    labels = scales::label_date_short()
    )+
  
  scale_fill_brewer(type = "qual")+
  
  labs(title = "COVID-19 epidemic, Fulton County, GA",
       subtitle = "Weekly reported cases among top five cities", 
       x = "Date of report",
       y = "Number of cases",
       fill = "City",
       caption = str_glue("Fake data for course exercise.\n{nrow(linelist)} cases shown by city of residence\nCases reported from {min(linelist$date_report)} to {max(linelist$date_report)}")) + 
  
    theme_minimal()+
     
    theme(legend.position = "right")
  


```


\pagebreak 

# Cumulative case incidence by city, per 10,000 population  

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
/// zip_counts \\\
--------------------------------------------------------------------------------

Here we use {dplyr} functions to get case counts based on ZIP Code and the 
time periods created in the define_reporting_periods code chunk above. 
We store as an object called zip_counts so that it can be merged with population
in the chunks below. 

For more details see: 
https://epirhandbook.com/descriptive-tables.html#dplyr-package
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

```{r zip_counts}
city_table <- linelist %>% 
  
  group_by(city) %>% 
  
  # create new summary columns 
  summarise(
    total_cases  = n(),
    recent_cases = sum(date_report > surveillance_date - 14)) %>% 

  # join with zip code populations data frame
  left_join(city_pops, by = c("city")) %>% 
   
  # add cululative case incidence  
  mutate(inc_per_10k = round((total_cases / population)* 10000, digits = 1)) %>% 

  # remove extra column
  select(-population) %>% 

  # sort on incidence per 10,000 population
  arrange(desc(inc_per_10k)) 
  
```


```{r}
# print table nicely and adjust the header labels  
city_table %>% 
  qflextable %>%  
  set_header_labels(         
      city = "City", 
      total_cases = "Total cases",                  
      recent_cases = "Cases in last 14 days",
      inc_per_10k = "Cumulative incidence per 10,000") 
```




\pagebreak 


## Table of Race and hospitalization (Yes/No)  

```{r}
# Compare race and died
linelist %>% 
     select(race, hospitalized) %>% 
     tbl_summary(by = hospitalized, percent = "row") %>% 
     add_p()
```


